---
title: "Cost Calculations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cost Calculations}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
```

The purpose of this vignette is to present the calculations of the costs for various distribution.

In the follow variables identified by Greek letters are considered known *a priori*.

# Univariate Guassian

Data belongs to group $k$ whose time stamps are the set $T_{k}$ can have additive mean anomaly $m_{k}$ and multiplicative variance anomaly $s_{k}$ which are common for $t \in T_{k}$.
For $t \in T_{k}$ this gives
\[
P\left(y_t \left| \mu_t,m_k,\sigma_k,s_k\right.\right) = \frac{1}{\sqrt{2\pi\sigma_{t}s_{k}}}\exp\left(-\frac{1}{2\sigma_{t}s_{k}}\left(y_{t} - \mu_t - m_{k}\right)^2\right)
\]

with the likelihood of of teh $n_{k}$ observations $y_{t \in T_{k}}$ being
\[
L\left(y_{t \in T_{k}} \left| \mu_t,m_k,\sigma_k,s_k\right.\right) = 
\left(2\pi s_{k}\right)^{-n_{k}/2} 
\prod\limits_{t \in T_{k}} \sigma_{t}^{-1/2}
\exp\left(-\frac{1}{2s_{k}}\sum\limits_{t \in T_{k}} \frac{\left(y_{t} - \mu_t - m_{k}\right)^2}{\sigma_{t}}\right)
\]

The log-likelihood of $y_{t \in T_{k}}$ is then 
\[
l\left(y_{t \in T_{k}} \left| \mu_t,m_k,\sigma_k,s_k\right.\right) = 
-\frac{n_{k}}{2} \log\left(2\pi s_{k}\right)
-\frac{1}{2}\sum\limits_{t \in T_{k}} \log\left(\sigma_{t}\right)
-\frac{1}{2s_{k}}\sum\limits_{t \in T_{k}} \frac{\left(y_{t} - \mu_t - m_{k}\right)^2}{\sigma_{t}}
\]

with the cost being twice the negative log likelihood plus a penalty $\beta$ giving 

\[
C\left(y_{t \in T_{k}} \left| \mu_t,m_k,\sigma_k,s_k\right.\right) = 
n_{k} \log\left(2\pi s_{k}\right)
+\sum\limits_{t \in T_{k}} \log\left(\sigma_{t}\right)
+\frac{1}{s_{k}}\sum\limits_{t \in T_{k}} \frac{\left(y_{t} - \mu_t - m_{k}\right)^2}{\sigma_{t}} + \beta
\]

## Anomaly in mean and varinace

Estimates $\hat{m}$ of $m$ and $\hat{\sigma}$ of $\sigma$ can be selected to minimise the cost by taking
\[
\hat{m}_{k} = \left( \sum\limits_{t \in T_k} \frac{y_t-\mu_t}{\sigma_t} \right)\left(  \sum\limits_{t \in T_k} \frac{1}{\sigma_t}\right)^{-1}
\]
and
\[
\hat{s}_{k} = \frac{1}{n_{k}} \sum\limits_{t \in T_k} \frac{ \left(y_t-\mu_t - \hat{m}_{k}\right)^2}{\sigma_t}
\]

Subsituting these into the cost gives
\[
C_{MV}\left(y_{t \in T_{k}} \left| \mu_t,\hat{m}_k,\sigma_k,\hat{s}_k\right.\right) = 
n_{k} \log\left(2\pi \hat{s}_{k}\right)
+\sum\limits_{t \in T_{k}} \log\left(\sigma_{t}\right)
+n_{k} + \beta
\]

## Anomaly in Mean

There is no change in variance so $s_{k}=1$. Estimate of $\hat{m}_{k}$ is unchanged from that for an anomaly in mean and variance so the cost

\[
C_{M}\left(y_{t \in T_{k}} \left| \mu_t,\hat{m}_k,\sigma_k\right.\right) = 
n_{k} \log\left(2\pi\right)
+\sum\limits_{t \in T_{k}} \log\left(\sigma_{t}\right)
+\sum\limits_{t \in T_k} \frac{ \left(y_t-\mu_t - \hat{m}_{k}\right)^2}{\sigma_t}
+ \beta
\]

can be written as

\[
C_{M}\left(y_{t \in T_{k}} \left| \mu_t,\hat{m}_k,\sigma_k\right.\right) = 
n_{k} \log\left(2\pi\right)
+\sum\limits_{t \in T_{k}} \log\left(\sigma_{t}\right)
+\sum\limits_{t \in T_k} \frac{ \left(y_t-\mu_t\right)^2}{\sigma_t}
-\hat{m}^{2} \sum\limits_{t \in T_k} \frac{ 1}{\sigma_t}
+ \beta
\]

## Anomaly in Variance

These is no mean anomaly so $m_{k}=0$. Estimate of $\hat{s}_{k}$ therfore changes to

\[
\hat{s}_{k} = \frac{1}{n_{k}} \sum\limits_{t \in T_k} \frac{ \left(y_t-\mu_t\right)^2}{\sigma_t}
\]
and cost is
\[
C_{V}\left(y_{t \in T_{k}} \left| \mu_t,\sigma_k,\hat{s}_k\right.\right) = 
n_{k} \log\left(2\pi \hat{s}_{k}\right)
+\sum\limits_{t \in T_{k}} \log\left(\sigma_{t}\right)
+n_{k}
+ \beta
\]

## No Anomaly (Baseline)

Here $m_{k}=0$ and $s_{k}=1$ and there is no penalty so

\[
C_{B}\left(y_{t \in T_{k}} \left| \mu_t,\sigma_t\right.\right) = 
n_{k} \log\left(2\pi \right)
+\sum\limits_{t \in T_{k}} \log\left(\sigma_{t}\right)
+\sum\limits_{t \in T_{k}} \frac{\left(y_{t} - \mu_t\right)^2}{\sigma_{t}} 
\]


## Point anomaly

A point anomaly at time $t$ is treated as a sinle time step with a variance anomaly.
Naively the cost could be computed using the formulea for a variance anomaly as
\[
C_{p}\left(y_{t}\left| \mu_{t},\sigma_{t}\right.\right) = \log\left(2\pi \frac{ \left(y_t-\mu_t\right)^2}{\sigma_t}\right)
+ \log\left(\sigma_{t}\right)
+ 1 + \beta
= \log\left(2\pi\right) + \log \left( \left(y_t-\mu_t\right)^2 \right) + 1 + \beta
\]

However the cost of the point anomaly should be higher then the background cost when $y_{t}$ is, in some sense, close to the background.

Follow Fisch et al. in intorducing a term $\gamma$ to control this.
Using the standardised variable $z_{t} = \frac{y_t-\mu_t}{\sqrt{\sigma_{t}}}$ the modified cost of a point anomaly is expressed as
\[
C_{p}\left(y_{t}\left| \mu_{t},\sigma_{t}\right.\right) = 
\log\left(2\pi\right) + \log\left(\sigma_{t}\right) + \log \left( \gamma + z_{t}^{2} \right) + 1 + \beta
\]

Relating this to the background cost we see that point anomalies may be accepted in the capa search when
\[
f\left(z_{t}^2,\gamma,\beta\right) = C_{p}\left(y_{t}\left| \mu_{t},\sigma_{t}\right.\right) - C_{B}\left(y_{t}\left| \mu_{t},\sigma_{t}\right.\right) = \log \left( \gamma + z_{t}^{2} \right) + 1 + \beta - z_{t}^{2} < 0
\]

This implies that $\gamma$ should be selected such that $f\left(0,\gamma,\beta\right) \geq 0$ 

The gradient wrt $z_{t}^2$ is
\[
\frac{\partial}{\partial z_{t}^2} f\left(z_{t}^2,\gamma,\beta\right)
= \frac{1}{\gamma + z_{t}^{2}} - 1
\]

Requiring $\gamma < 1$; which maintains a positive gradient for small z_{t}^2; indicates that there will no anomalies near 0.

Consider three different definitions of $\gamma$.

- The non corection of $\gamma_{0} = 0$ which allows point anomalies as z_{t}^2 approaches 0
- The correction $\gamma_{1} = \exp\left(-\beta\right)$ proposed by Fisch et al.
- The minimal correction $\gamma_{2} = \exp\left(-\left(1+\beta\right)\right)$ for which $f\left(0,\gamma_{2},\beta\right) = 0$.

To see the impact of the correct for small $z$ the following figure shows 
$f\left(z_{t}^2,\gamma,\beta\right)$ for $\beta=10$ for the three options.

```{r echo=FALSE, f, fig.width=7, fig.height=7}
zsq <- seq(0,1e-3,length=10000)
Y <- matrix(NA,length(zsq),3)
colnames(Y) <- c("gamma_0","gamma_1","gamma_2")
b <- 10
Y[,1] <- fz(zsq,0,b)
Y[,2] <- fz(zsq,exp(-b),10)
Y[,3] <- fz(zsq,exp(-(1+b)),10)


matplot(sqrt(zsq),Y,type="l",xlab="z",ylab="f(z,gamma,10)");
legend("bottomright",colnames(Y),col=1:3,lty=1:3)
```

It is clear that the difference become small as $z$ increases. This is supported by the plot below shows the value of $z_{t}$ at which an point anomaly might occur as $\beta$ varies. Area above the line are potential anomaly values.

```{r echo=FALSE, when_anaom, fig.width=7, fig.height=7}
fz <- function(zsq,gamma,beta){ log(gamma+zsq) + 1 + beta - zsq }
betaRng <- seq(log(2),2*log(10),length=1000) ###2*log(1:1000)
Y <- matrix(NA,length(betaRng),3)
colnames(Y) <- c("gamma_0","gamma_1","gamma_2")

for(ii in 1:length(betaRng)){
    rng <- c(1.25,25)
    b <- betaRng[ii]
    Y[ii,1] <- uniroot(fz,rng,gamma=0,beta=b)$root
    Y[ii,2] <- uniroot(fz,rng,gamma=exp(-b),beta=b)$root
    Y[ii,3] <- uniroot(fz,rng,gamma=exp(-(1+b)),beta=b)$root
}
matplot(betaRng,sqrt(Y),type="l",xlab="beta",ylab="z");
legend("bottomright",colnames(Y),col=1:3,lty=1:3)
```
