---
title: "Univariate Poisson Cost Calculations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Univariate Poisson Cost Calculations}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
```

The purpose of this vignette is to present the calculations of the costs for the univariate Poisson distribution.

Each time step $t$ belongs to group $k$ whose time stamps are the set $T_{k}$. A group can have a multiplicative rate anomaly $\lambda_{k}$ which is common for $t \in T_{k}$. Assuming the {\it a priori} known rate $r_{t}$ the data generating distribution gives for $t \in T_{k}$

\[
P\left(y_t \left| r_t, \lambda_k\right.\right) = 
\frac{\lambda_{k}^{y_{t}} r_{t}^{y_{t}} \exp\left(-r_{t}\lambda_{k}\right)}{y_{t}!}
\]


The cost is computed as twice the negative log likelhiood plus a penalty term $\beta$ giving

\[
C\left(y_{t \in T_{k}} \left| \lambda_k,r_{t \in T_{k}}\right.\right) = 
2 \lambda_{k} \sum\limits_{t \in T_{k}} r_{t}
- 2 \log\left(\lambda_{k}\right) \sum\limits_{t \in T_{k}} y_{t} 
- 2 \sum\limits_{t \in T_{k}} y_{t} \log\left(r_{t}\right)
+ 2 \sum\limits_{t \in T_{k}} \log\left( y_{t}! \right)
+ \beta
\]


### No Anomaly (Baseline)

In this case $\lambda_{k}=1$ and there is no penalty so

\[
C_{B}\left(y_{t \in T_{k}} \left| r_{t \in T_{k}}\right.\right) = 
2 \sum\limits_{t \in T_{k}} r_{t}
- 2 \sum\limits_{t \in T_{k}} y_{t} \log\left(r_{t}\right)
+ 2 \sum\limits_{t \in T_{k}} \log\left( y_{t}! \right)
\]

### Anomaly in Rate

An estimate $\hat{\lambda}_{k}$ of $\lambda_{k}$ can be selected to minimise the cost by taking
\[
\hat{\lambda}_{k} = \frac{ \sum\limits_{t \in T_{k}} y_{t} }{\sum\limits_{t \in T_{k}} r_{t}}
\]

\[
C_{A}\left(y_{t \in T_{k}} \left| \hat{\lambda}_k,r_{t \in T_{k}}\right.\right) = 
2 \sum\limits_{t \in T_{k}} y_{t}
- 2 \log\left(\hat{\lambda}_{k}\right) \sum\limits_{t \in T_{k}} y_{t} 
- 2 \sum\limits_{t \in T_{k}} y_{t} \log\left(r_{t}\right)
+ 2 \sum\limits_{t \in T_{k}} \log\left( y_{t}! \right)
+ \beta
\]

An anomaly will be accepted whenever

\[
C_{A}\left(y_{t \in T_{k}} \left| \hat{\lambda}_k,r_{t \in T_{k}}\right.\right) - 
C_{B}\left(y_{t \in T_{k}} \left| r_{t \in T_{k}}\right.\right) = 
2 \sum\limits_{t \in T_{k}} y_{t}
- 2 \log\left(\hat{\lambda}_{k}\right) \sum\limits_{t \in T_{k}} y_{t} 
+ \beta
- 2 \sum\limits_{t \in T_{k}} r_{t}
<0
\]

Rearranging this expression in terms of $\sum\limits_{t \in T_{k}} r_{t}$, the expected number of counts if the period was not anomalous gives

\[
C_{A}\left(y_{t \in T_{k}} \left| \hat{\lambda}_k,r_{t \in T_{k}}\right.\right) - 
C_{B}\left(y_{t \in T_{k}} \left| r_{t \in T_{k}}\right.\right) = 
\beta -
2 \left( 1 - \hat{\lambda}_{k} + \hat{\lambda}_{k}\log\left(\hat{\lambda}_{k}\right)\right)
\sum\limits_{t \in T_{k}} r_{t}
\]

This form suggests the selection of $\beta$ based on a minimum change in $\hat{\lambda}_{k}$ away from 1 causing at least a certain change from the expected number of counts.
The figure below show $\gamma = 2 \left( 1 - \hat{\lambda}_{k} + \hat{\lambda}_{k}\log\left(\hat{\lambda}_{k}\right)\right)$. Using this we could select $gamma=0.01$ to ensure a change in in lambda of 10\%, which combined with a desire to detect only changes of 50 units would result in $\beta = 0.5$

```{r echo=FALSE, f, fig.width=7, fig.height=7}
lambda <- seq(0.7,1.3,length=100000)
gamma <- 2*(1-lambda+lambda*log(lambda))
plot(lambda,gamma,type="l",xlab="lambda",ylab="gamma")
```


